Exploring dimensions of biodiversity in Japanese ferns

### 2021-07-01 ms/manuscript.Rmd
Add t-value and model fit plots to ms

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro"
))
```

@@

All spatial models were significant relative to null models without any fixed effects. The effect of temperature dominated in most models, with only SES of PD showing a stronger effect of precipitation seasonality (`r figure("coeff")`A--E). Biodiversity metrics with a genetic component (SES of PD and SES of RPD) each had a large, negative relationship with % apomictic taxa (`r figure("coeff")`F--G). Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD.

@@

# Format parameter values for plotting
model_params_for_plot <-
  model_params %>%
  # add model type
  tidyr::extract(formula, "model_type", regex = " ~ ([^ ]+) ") %>%
  # drop intercept
  filter(term != "(Intercept)") %>%
  # add LRT results and names formatted for printing
  left_join(lrt_signif, by = c("resp_var", "term", "model_type")) %>%
  left_join(resp_var_lookup, by = "resp_var") %>%
  left_join(indep_var_lookup, by = "term") %>%
  left_join(model_type_lookup, by = "model_type") %>%
  # Make coefficient with greatest abs value bold
  group_by(resp_var, model_type) %>%
  mutate(
    term_print = case_when(
      abs(estimate) == max(abs(estimate)) ~ glue::glue("**{term_print}**"),
      TRUE ~ term_print
    ),
    term_print = factor(term_print)
  ) %>%
  ungroup()

@@

  ggplot(plot_data, aes(x = estimate, y = term_print, color = signif)) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    geom_pointrange(aes(xmin = estimate - cond_se, xmax = estimate + cond_se), size = 0.25) +
    scale_x_continuous(n.breaks = 4) +
    scale_color_manual(
      values = c("TRUE" = "grey20", "FALSE" = "grey70"),
      labels = c("TRUE" = "*p* < 0.05", "FALSE" = "*p* > 0.05")) +
    labs(
      subtitle = glue("{unique(plot_data$resp_var_print)},\n{unique(plot_data$model_type_print)}"),
      color = "Significance") +
    theme_gray(base_size = 9) +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_markdown(),
      legend.text = element_markdown(),
      legend.position = "none")
}

@@

**`r figure("coeff")`**. Regression coefficients of general mixed models. A--E, environmental models (models including temperature); F and G, reproductive models (models including % apomictic taxa). Error bars show 95% confidence interval. Statistical significance assessed with a likelihood ratio test (LRT) between the full model and a model with the focal independent variable removed (null model); *p*-value indicates probability of no difference between the full and null model. Name of response variable and model type indicated above each subplot. Independent variable with greatest absolute coefficient value in bold for each model.

>>>>

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  # "plastome_calibration_dates",
  # FIXME: remove until phylo analysis is working "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro",
  "model_fits",
  "spatial_models"
))
```

@@

All spatial models were significant relative to null models without any fixed effects. The effect of temperature dominated in environmental models (`r figure("coeff")`A--E), and % apomictic taxa dominated in reproductive models (`r figure("coeff")`F--G). Richness showed a hump-shaped relationship with temperature, reaching maximum richness at intermediate temperatures (`r figure("modelfit")`A). Phylogenetic and functional diversity showed either moderately (SES of PD, SES of FD) to strongly (SES of RPD, SES of RFD) hump-shaped relationships with temperature (`r figure("modelfit")`B--E). SES of RPD showed a clear, linear decline with % apomictic taxa (`r figure("modelfit")`G); a similar pattern was detected in SES of PD, but with wider confidence intervals (`r figure("modelfit")`F). Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD.

@@

# Format parameter values for plotting
model_params_for_plot <-
  model_params %>%
  # add model type
  tidyr::extract(formula, "model_type", regex = " ~ ([^ ]+) ") %>%
  # drop intercept
  filter(term != "(Intercept)") %>%
  # add LRT results and names formatted for printing
  left_join(lrt_signif, by = c("resp_var", "term", "model_type")) %>%
  left_join(resp_var_lookup, by = "resp_var") %>%
  left_join(indep_var_lookup, by = "term") %>%
  left_join(model_type_lookup, by = "model_type") %>%
  # Make coefficient with greatest abs t-value bold
  group_by(resp_var, model_type) %>%
  mutate(
    term_print = case_when(
      abs(t_value) == max(abs(t_value)) ~ glue::glue("**{term_print}**"),
      TRUE ~ term_print
    ),
    term_print = factor(term_print)
  ) %>%
  ungroup()

@@

  ggplot(plot_data, aes(x = t_value, y = term_print, fill = signif)) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    geom_col() +
    # scale_x_continuous(n.breaks = 4) +
    scale_x_continuous() +
    scale_fill_manual(
      values = c("TRUE" = "grey20", "FALSE" = "grey70"),
      labels = c("TRUE" = "*p* < 0.05", "FALSE" = "*p* > 0.05")) +
    labs(
      subtitle = glue("{unique(plot_data$resp_var_print)},\n{unique(plot_data$model_type_print)}"),
      fill = "Significance") +
    theme_gray(base_size = 9) +
    theme(
      axis.title = element_blank(),
      axis.text.y = element_markdown(),
      legend.text = element_markdown(),
      legend.position = "none")
}

@@

**`r figure("coeff")`**. Regression coefficient *t*-values (coefficient divided by standard error) of general mixed models. A--E, environmental models (models including temperature); F and G, reproductive models (models including % apomictic taxa). Statistical significance assessed with a likelihood ratio test (LRT) between the full model and a model with the focal independent variable removed (null model); *p*-value indicates probability of no difference between the full and null model. Name of response variable and model type indicated above each subplot. Independent variable with greatest absolute *t*-value in bold for each model.

```{r modelfit-make}
# Make plot showing model fits

# Extract dataset used for model, including significance columns
biodiv_ferns_used_for_model <- sf_to_centroids(biodiv_ferns_spatial) %>%
  # filter to only sites used in model
  inner_join(
    spatial_models$model[[1]]$data %>% as_tibble() %>% select(grids),
    by = "grids"
  )

# Specify order of subplots
model_subplot_order <- tribble(
  ~resp_var, ~indep_var, ~order,
  "richness", "temp", 1,
  "pd_obs_z", "temp", 2,
  "rpd_obs_z", "temp", 3,
  "fd_obs_z", "temp", 4,
  "rfd_obs_z", "temp", 5,
  "pd_obs_z", "percent_apo", 6,
  "rpd_obs_z", "percent_apo", 7
)

# Define function for plotting model fit and actual data
plot_fits <- function(resp_var, resp_var_print, indep_var, signif_var, fit, model_data, ...) {

  # Setup randomization test significance colors
  signif_cols <-
    c(
      "> 0.99" = "#e31a1c", 
      "> 0.975" = "#fb9a99",
      "< 0.01" = "#1f78b4", 
      "< 0.025" = "#a6cee3",
      "not significant" = "#ffffcc"
    )

  # Basic plot setup
  p <- ggplot(model_data, aes(x = .data[[indep_var]], y = .data[[resp_var]])) +
    labs(y = resp_var_print) +
    theme(panel.grid.minor = element_blank())

  # Add points, color by significance if not plotting richness
  if (resp_var != "richness") {
    p <- p + 
      geom_point(aes(color = .data[[signif_var]]), size = 0.5) +
      scale_color_manual(values = signif_cols)
  } else {
    p <- p + 
      geom_point(size = 0.5, color = "grey20")
  }

  # Add line and ribbon of model fit
  p <- p +
    geom_line(data = fit, alpha = 0.35, color = "blue") +
    geom_ribbon(
      data = fit,
      aes(ymin = low, ymax = up), 
      alpha = 0.15
    ) +
    theme(legend.position = "none")

  # Format x-axis ticks and labels
  if (indep_var == "temp") {
    p <- p +
      scale_x_continuous(labels = function(x) x*0.1) +
      labs(x = "Temperature (°C)")
  } else if (indep_var == "percent_apo") {
    p <- p +
      scale_x_continuous(labels = function(x) scales::percent(x, accuracy = 1)) +
      labs(x = "% apomictic taxa")
  }

  p

}

# Generate final plot
modelfit_plot_list <- 
  model_fits %>%
  mutate(
    signif_var = str_replace_all(resp_var, "_obs_z", "_signif"),
    resp_var_print = case_when(
      resp_var == "richness" ~ "Richness (no. spp.)",
      TRUE ~ str_to_upper(resp_var) %>% word(1, 1, "_") %>% paste("SES of", .)
    ),
    model_data = list(biodiv_ferns_used_for_model)) %>%
  left_join(model_subplot_order, by = c("resp_var", "indep_var")) %>%
  arrange(order) %>%
  pmap(plot_fits)

# Need one subplot with a legend
modelfit_plot_list[[2]] <- modelfit_plot_list[[2]] +
  theme(legend.position = "right") +
  labs(color = "Significance") +
  # Increase size of points in legend
  guides(colour = guide_legend(override.aes = list(size = 1.5)))

modelfit_plot <-
  wrap_plots(modelfit_plot_list) +
  guide_area() +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

```{r modelfit-save, eval = params$doc_type == "doc"}
# Save plot if rendering to doc
ggsave(result_file(figure("modelfit"), "pdf"), modelfit_plot, width = 7, height = 5, units = "in")
```

```{r modelfit-print, fig.width = 7, fig.height = 5, eval = params$doc_type == "pdf"}
# Print plot if rendering to PDF
modelfit_plot
```

**`r figure("modelfit")`**. Relationship between biodiversity metrics and selected predictor variables. A--E, environmental models (models including temperature); F and G, reproductive models (models including % apomictic taxa). Ribbon shows 95% confidence interval of model. Line fit with the focal predictor variable while averaging over other predictors. For biodiversity metrics other than richness, raw values were compared to a null distribution of 999 random values, and the *P*-value calculated using a two-tailed test.
